Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-10 paper/paper.tex
intro/elim/beta/eta for indexed inductives on paper, and start agda formalization of on-paper rules
remove unused fig
some syntax

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}

@@

@@

@@

    \inferrule*[right=IndElim]
    {}
    {\Gamma; \Delta \vdash \mathsf{fold}(x.b.f)(M, e) : B\,M}
  \end{mathpar}
  \caption{Indexed Inductive Linear Types}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \begin{array}{rcl}
      \textrm{non-linear types} X,Y,Z & ::= & \textrm{Empty}\pipe\textrm{Unit}\pipe\textrm{Bool}\pipe\textrm{Nat}
      \pipe \sum_{x:X} Y \pipe \prod_{x:X}Y\pipe \textrm{SPF}\,X \pipe \cdots \max{also eliminators...}\\
      \textrm{linear types} A,B,C & ::= & c \pipe I \pipe A \otimes B \pipe A \lto B \pipe B \tol A
      \pipe \bigoplus_{x:X} A \pipe \bigamp_{x:X} A \pipe \mu e e'\\
      \textrm{non-linear terms} M,N,O & ::= & \cdots\\
      \textrm{linear terms} 
    \end{array}
  \end{mathpar}
  \caption{Syntax of Types and Terms}
\end{figure}

@@

  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \and
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \and
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \and
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
  \end{mathpar}
  \caption{Intuitionistic typing}
  \label{fig:inttyping}
\end{figure}

>>>>

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}

@@

\newcommand{\equalizer}[4]{\{#1\,|\,\apptol {#2}{#1} = \apptol{#3}{#1 \}}

@@

\begin{figure}
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    % 
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe \mathsf{Int} \pipe \mathsf{Empty} \pipe \mathsf{Unit} \pipe \mathsf{SPF}\,X\pipe U \pipe L \pipe !A \pipe \sum_{x:X} Y \pipe \prod_{x:X} Y\pipe M =_A N\\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{e}{f}\\
  \end{array}
  \caption{Overview of Types}
\end{figure}

@@

    \inferrule*[right=IndElim]
    {\Gamma;\Delta \vdash e : \mu A M\and
     \Gamma,x:X;b:\el(A\,x)\,B \vdash f : B\,x
    }
    {\Gamma; \Delta \vdash \mathsf{fold}(A)(x.b.f)(M, e) : B\,M}

    \inferrule*[right=Ind$\beta$]
    {}
    {\fold(A)(x.b.f)(M, \roll(e)) =
      f[M/x][\map(A\,M)\,(y.a.\fold(A)(x.b.f)(y,a))\,e/b]}

    \inferrule*[right=Ind$\eta$]
    {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
      \and \Gamma;\Delta \vdash e : \mu A M\\\\
      \Gamma,x:X; a':\el(A x)(\mu A) \vdash
      e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
    }
    {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
  \end{mathpar}
  \caption{Indexed Inductive Linear Types}
\end{figure}

@@

  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \and
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \and
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \and
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}

  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
  \end{mathpar}
  \caption{Intuitionistic typing}
  \label{fig:inttyping}
\end{figure}
